Type Refinements for the Working Class

### 2015-12-24 paper.tex
Add critical comparison w/ Allen's semantics (#9)
@freebroccolo This is kind of interesting---I was thinking about the
internal observations that can be made in our semantics but which cannot
under Allen's, and now that I think of it, I believe that the
open-endedness of the supervaluation semantics relies on a condition
which is actually impossible if you take a Brouwerian view concerning
construction and data.

\subsubsection{The type of atoms}
\newcommand\SortAtom{\mathtt{atom}}
\newcommand\OpToken[1]{\mathtt{tok}[#1]}
\newcommand\OpMatch[1]{\mathtt{match}_{#1}}

@@

Then, we can define a refinement $\GloballyRefines{\mathbb{A}_0}{\SortAtom}$ as
follows:
\[
  \infer{
    \RMemEq{\Upsilon}{
      \App{\OpToken{a}}{}
    }{
      \App{\OpToken{a}}{}
    }{\mathbb{A}_0}
  }{
    \IMode{\Upsilon}\ni\IMode{a}:\IMode{\SortAtom}
  }
\]

However, it is not the case that $\mathbb{A}_0$ is a type, because it doesn't
respect computational equivalence. This is easily resolved by defining the type
of atoms as the computational extension of $\mathbb{A}_0$,
\[
  \Define{
    \mathbb{A}
  }{
    \VToCType(\mathbb{A}_0)
  }
\]

\begin{thm}
  For any $\Upsilon$, we have $\IsType{\Upsilon}{\mathbb{A}}{\SortAtom}$.
\end{thm}
\begin{proof}
  Trivial, because the computational extension of any value PER is a type.
\end{proof}

@@

For instance, assuming a type $\Naturals$ of natural numbers and $\mathbb{A}$
of atoms, with $\Naturals\inj\mathbb{A}$ the type of embeddings of the naturals
into the atoms, we do have
$\NuprlInterpSeq{\Strings}{\cdot}{\Naturals\inj\mathbb{A}}$, because there are
numerous ways to embed the naturals into the strings; however, because the
embedding fails in every finite model $\AtomModel$, we do not have
$\NuprlInterpSeq{*}{\cdot}{\Naturals\inj\mathbb{A}}$.  Furthermore, we also do
not have $\NuprlInterpSeq{*}{\iota:\Naturals\inj\mathbb{A}}{\bot}$, because this
would fail in any infinite model.

@@

\begin{thm}
  \label{thm:no-nat-injection}
  As in the supervaluation semantics, we have
  $\not\models\NuprlSeq{\cdot}{\Naturals\inj\mathbb{A}}$.
\end{thm}
\begin{proof}
  Suppose $\models\NuprlSeq{\cdot}{\Naturals\inj\mathbb{A}}$. Then, there
  exists an $\Member{\Upsilon}{\SCtx}$ and $\Member{M}{\Exprs_\tau(\Upsilon)}$
  such that $\RMemEq{\Upsilon}{M}{M}{\Naturals\inj\mathbb{A}}$; but this
  is impossible because it would require a means of embedding the naturals into
  finite $\Dom{\Upsilon}$.
\end{proof}

\begin{thm}
  Contrary to the supervaluation semantics, we do have
  $\models\NuprlSeq{\iota:\Naturals\inj\mathbb{A}}{\bot}$.
\end{thm}
\begin{proof}
  It suffices to show that for all $\Upsilon$, there is no $\iota$ such that
  $\RMemEq{\Upsilon}{\iota}{\iota}{\Naturals\inj\mathbb{A}}$. However, by the
  definition of $\SCtx$, all $\Member{\Upsilon}{\SCtx}$ are finite, so there
  can be no injection of the naturals into $\Dom{\Upsilon}$.
\end{proof}

The crux of the problem is that all worlds (symbol contexts) are finite in our
semantics, so it is possible to observe that it will never be possible to
inject the naturals into the atoms. One way to make our system more open-ended
is to replace the current semantic universe $\Define{\SemU}{\Classes^\SCtx}$
with an extended universe $\Define{\SemU_\infty}{\Classes^{\SCtx_\infty}}$ with
$\SCtx_\infty$ the set of possibly-infinite symbol contexts and injective,
sort-preserving renamings. Externally, by Theorem~\ref{thm:no-nat-injection} we
could still never exhibit an embedding of the naturals, but internally, we
would be unable to rule it out.

>>>>

\subsubsection{The type of atoms}
\newcommand\SortAtom{\mathtt{atom}}
\newcommand\OpToken[1]{\mathtt{tok}[#1]}
\newcommand\OpMatch[1]{\mathtt{match}_{#1}}
\newcommand\Atoms{\mathbb{A}}

@@

Then, we can define a refinement $\GloballyRefines{\Atoms_0}{\SortAtom}$ as
follows:
\[
  \infer{
    \RMemEq{\Upsilon}{
      \App{\OpToken{a}}{}
    }{
      \App{\OpToken{a}}{}
    }{\Atoms_0}
  }{
    \IMode{\Upsilon}\ni\IMode{a}:\IMode{\SortAtom}
  }
\]

However, it is not the case that $\Atoms_0$ is a type, because it doesn't
respect computational equivalence. This is easily resolved by defining the type
of atoms as the computational extension of $\Atoms_0$,
\[
  \Define{
    \Atoms
  }{
    \VToCType(\Atoms_0)
  }
\]

\begin{thm}
  For any $\Upsilon$, we have $\IsType{\Upsilon}{\Atoms}{\SortAtom}$.
\end{thm}
\begin{proof}
  Trivial, because the computational extension of any value PER is a type.
\end{proof}

@@

For instance, assuming a type $\Naturals$ of natural numbers and $\Atoms$
of atoms, with $\Naturals\inj\Atoms$ the type of embeddings of the naturals
into the atoms, we do have
$\NuprlInterpSeq{\Strings}{\cdot}{\Naturals\inj\Atoms}$, because there are
numerous ways to embed the naturals into the strings; however, because the
embedding fails in every finite model $\AtomModel$, we do not have
$\NuprlInterpSeq{*}{\cdot}{\Naturals\inj\Atoms}$.  Furthermore, we also do
not have $\NuprlInterpSeq{*}{\iota:\Naturals\inj\Atoms}{\bot}$, because this
would fail in any infinite model.

@@

\begin{thm}
  \label{thm:no-nat-injection}
  As in the supervaluation semantics, we have
  $\not\models\NuprlSeq{\cdot}{\Naturals\inj\Atoms}$.
\end{thm}
\begin{proof}
  Suppose $\models\NuprlSeq{\cdot}{\Naturals\inj\Atoms}$. Then, there
  exists an $\Member{\Upsilon}{\SCtx}$ and $\Member{M}{\Exprs_\tau(\Upsilon)}$
  such that $\RMemEq{\Upsilon}{M}{M}{\Naturals\inj\Atoms}$; but this
  is impossible because it would require a means of embedding the naturals into
  finite $\Dom{\Upsilon}$.
\end{proof}

\begin{thm}
  \label{thm:never-nat-injection}
  Contrary to the supervaluation semantics, we do have
  $\models\NuprlSeq{\iota:\Naturals\inj\Atoms}{\bot}$.
\end{thm}
\begin{proof}
  It suffices to show that for all $\Upsilon$, there is no $\iota$ such that
  $\RMemEq{\Upsilon}{\iota}{\iota}{\Naturals\inj\Atoms}$. However, by the
  definition of $\SCtx$, all $\Member{\Upsilon}{\SCtx}$ are finite, so there
  can be no injection of the naturals into $\Dom{\Upsilon}$.
\end{proof}

Another seemingly paradoxical observation is also possible:

\begin{thm}
  \label{thm:never-atom-injection}
  The atoms may not be embedded into the naturals, i.e.\
  $\models\NuprlSeq{\iota:\Atoms\inj\Naturals}{\bot}$.
\end{thm}
\begin{proof}

  Fix a symbol context $\Upsilon$; suppose we have an $\Member{\iota}{\Exprs(\Upsilon)}$ such that
  $\RMemEq{\Upsilon}{\iota}{\iota}{\Atoms\inj\Naturals}$. Then, by inversion we
  have
  $\FRMemEq{\Upsilon}{x:\Atoms}{\App{\OpAp{}{}}{\iota;x}}{\App{\OpAp{}{}}{\iota;x}}{\Naturals}$;
  in other words, for all
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
  and $\Member{a}{\Dom{\Upsilon}}$, we have
  $\RMemEq{\Upsilon'}{\App{\OpAp{}{}}{\iota;a}}{\App{\OpAp{}{}}{\iota;a}}{\Naturals}$
  such that $\iota$ is injective.
  But the only way for $\iota$ to assign a different index to each atom is for
  it to test each atom in $\Dom{\Upsilon'}$ separately---but $\iota$ may only
  mention the atoms in $\Dom{\Upsilon}$.
\end{proof}

\paragraph{Should the above be worrisome?}

At the very least, the fact that neither the naturals nor the atoms may be
embedded into the other commits us to a truly intuitionistic ontology, as
opposed to merely constructive one. In particular, our framework is
anti-classical in at least a na\"ive sense, because the Axiom of Choice puts
each set in bijection with a cardinal, whence it is impossible for neither of
two sets to embed in the other. However, it is important to note that as far as
the propositions-as-types principle is concerned, Computational Type Theory is
already anti-classical, as the undecidability of the halting program can be
observed internally.

In fact, we view the simultaneous validity of
Theorems~\ref{thm:never-nat-injection}~and~\ref{thm:never-atom-injection} as
totally desirable, because we adhere to a strictly Brouwerian conception of
data. The proper way to view the type of atoms is as an ongoing act of free
generation---and it is a basic fact concerning the activity of an idealized
mathematician that at each stage, only a finite number of constructions shall
have been effected so far.

On the other hand, (the analogues of) these theorems are \emph{not} valid under
Allen's supervaluation semantics, because one would have to account for worlds
at which an infinite number of atoms have already been generated; in this way,
the open-endedness of Allen's semantics relies crucially on a condition that is
simply impossible under a Brouwerian conception of mathematical activity.

\begin{remark}
  One way to recover something essentially equivalent to the supervaluation
  semantics in our style would be to replace the current semantic universe
  $\Define{\SemU}{\Classes^\SCtx}$ with an extended universe
  $\Define{\SemU_\infty}{\Classes^{\SCtx_\infty}}$, $\SCtx_\infty$ the set of
  possibly-infinite symbol contexts and injective, sort-preserving renamings.
  Externally, by Theorem~\ref{thm:no-nat-injection} we could still never
  exhibit an embedding of the naturals, but internally, we would be unable to
  rule it out.
\end{remark}
